---
title: "cde-output-reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cde-output-reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cde)
```

## Reference list of all output columns

The table below gives information on all the current columns of data retrieved by the different functions of `cde`; names, what they mean and which functions return them. The EA do modify the format of their data from time to time, so this will be updated as necessary.


Column name | Description | Output by which functions
--- | --- | ---
river_basin_district | Name of the River Basin District. This is the top level in the [catchment hierarchy](https://environment.data.gov.uk/catchment-planning/help#help-catchment-hierarchy). | All functions
management_catchment | Management Catchments are the next level down the hierarchy, see link above. | All functions
operational_catchment | Operational Catchments are the next level down from Management Catchments. | All functions
waterbody_id | Water Bodies are the smallest unit in the catchment hierarchy, representing all or part of a river system, lake, estuary etc. Each has a unique ID (WBID). | All functions
status | WFD status classification (High, Good, Moderate, Poor or Bad). For objectives this represents the target (aimed for) status class. | All get_ functions except get_pa
classification_level | The level within the range of classification elements that the status class refers to (NEEDS LINK) | objectives, rnag & status
classification_level |  | -
cycle |  | -
water_body |  | -
water_body_type |  | -
year |  | -
classification_item |  | -
easting |  | -
hydromorphological_designation |  | -
id |  | -
investigation_outcome |  | -
ngr |  | -
northing |  | -
activity_certainty |  | -
business_sector |  | -
category |  | -
category_certainty |  | -
certainty |  | -
classification_element |  | -
classification_id |  | -
confidence |  | -
estimated_start_date |  | -
funding_stream |  | -
lead_organisation |  | -
linked_rnags |  | -
measure_category_1 |  | -
measure_category_2 |  | -
measure_category_3 |  | -
measure_reference_code |  | -
measure_type |  | -
national_swmi_header |  | -
objectivetype |  | -
pressure_tier_1 |  | -
pressure_tier_2 |  | -
pressure_tier_3 |  | -
protected_area_code |  | -
protected_area_label |  | -
protected_area_type |  | -
reason_type |  | -
reasons_for_alternative_objectives |  | -
rnag_url |  | -
sector_of_lead_organisation |  | -
see_also |  | -
swmi |  | -
swmi_certainty |  | -
title |  | -

