---
title: "cde: Reference list of all output columns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cde-output-reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(cde)
```
The table below gives information on all the current columns of data retrieved by the different `get_...` functions of `cde`; names, what they mean and which functions return them. The EA do modify the format of their data from time to time, so this will be updated as necessary.

Column name | Description | Output from which functions
--- | --- | ---
river_basin_district | Name of the River Basin District. This is the top level in the [catchment hierarchy](https://environment.data.gov.uk/catchment-planning/help#help-catchment-hierarchy). | All get_... functions
management_catchment | Name of the Management Catchment, the next level down the hierarchy, see link above. | All get_... functions
operational_catchment | name of the Operational Catchment, the next level down from Management Catchments. | All get_... functions
waterbody_id | Waterbodies are the smallest unit in the catchment hierarchy, representing all or part of a river system, lake, estuary etc. Each has a unique ID (WBID). | All get_... functions
status | WFD status classification (High, Good, Moderate, Poor or Bad). For objectives this represents the target (aimed for) status class. | All get_... functions except get_pa
classification_item | Broad category of classification level (ecological, chemical or quantitative). | get_objectives, get_rnag & get_status
classification_level | The level within the range of classification elements that the status class refers to (see details [here](https://environment.data.gov.uk/catchment-planning/help#help-surface-waters-classification-hierarchy)). | get_objectives, get_rnag & get_status
cycle | Which cycle of assessment the data come from. The first cycle. The first cycle ran from 2009 to 2015 and cycle 2 to 2015-2021. There were changes in some environmental standards between cycles and for the overlap period, `cde` only returns cycle 2 assessments. | get_objectives, get_rnag & get_status
water_body | Name of the waterbody. | get_objectives, get_rnag & get_status
water_body_type | Type of waterbody. Values are: River, Lake, CoastalWater, TransitionalWater, GroundWaterBody | get_objectives, get_rnag & get_status
year | Year to which the data refer to. | get_objectives, get_rnag & get_status
easting | British National Grid Reference easting, i.e. x coordinate of location. | get_objectives & get_status
hydromorphological_designation | Hydromorphological designation of the waterbody (e.g. whether the body is artificial or heavily modified). | get_objectives & get_status
ngr | Ordnance Survey Landranger grid reference of the location. | get_objectives & get_status
northing | British National Grid Reference northing, i.e. y coordinate of location.  | get_objectives & get_status
activity | Classification of the activity linked to the RNAG identified, for example "Forestry", "Septic Tanks". | get_rnag
activity_certainty | Assessment of the certainty with which the RNAG identified is linked to the activity specifed.  | get_rnag
business_sector | Categorisation of the business sector linked to the activity causing the RNAG, for example "Transport", "Retail sector", "Forestry". | get_rnag
category | Broad categorisation of the RNAG, for example "Industry", "Agriculture and rural land management" or "Water Industry". | get_rnag
category_certainty | Assessment of the certainty with which the RNAG identified is linked to the category specifed. | get_rnag
certainty | Assessment of the certainty of the status class assignment. | get_status
classification_element | Specific quality element or metric linked to the RNAG. | get_rnag
confidence | Confidence in the status class assignment (0-1 scale). | get_status
estimated_start_date | Estimated date at which action specified will being  | get_measures
funding_stream | Source of funding for measure specified. | get_measures
id | Unique numerical code of RNAG (may be linked to measures specified).  | get_rnag
investigation_outcome | Outcome of the investigation into the RNAG | get_rnag
lead_organisation | Organisation leading the work on the measure specified. | get_measures
linked_rnags | Numerical code(s) of RNAGS linked to measures specified (see 'id' above.)  | get_measures
measure_category_1 | Top level classification of measure specified. | get_measures
measure_category_2 | Next level down (more detailed) classification of measure specified. | get_measures
measure_category_3 | Most detailed classification of measure specified. | get_measures
measure_reference_code | Numerical code of measure type. | get_measures
measure_type | Classification of type of measure employed. | get_measures
national_swmi_header | SWMI (Significant Water Management Issue) classification at the national level, for example "Pollution from waste water" or "Physical modifications". | get_rnag
objectivetype | Type of objective set, either "Predicted" or "Objective" | get_objectives
pressure_tier_1 | Detailed classification of the pressure related to the RNAG identified, for example "Irrigation", "Priority Substances (pesticides)". | get_rnag
pressure_tier_2 | Specific substance or pressure associated with RNAG identiifed, for example "Phosphate", "High temperature". | get_rnag
pressure_tier_3 | High level classificationb of the pressure related to the RNAG identifed, for example "Physical modification", "Dissolved oxygen (DO)". | get_rnag
protected_area_code | Unique code of each protected area. | get_pa
protected_area_label | Name or code of protected area. | get_pa
protected_area_type | Designation of protected area, for example "Nitrates Directive" ,  "Bathing Water Directive" | get_pa
reason_type | Type of RNAG, either "RFF" - reason for failure or "RFD" - reason for deterioration | get_rnag
reasons_for_alternative_objectives | Detailed text as to why alternative objectives i.e. a status less than good, have been set. | get_objectives
rnag_url |  URL to page on CDE site giving specifics of RNAG | get_rnag
sector_of_lead_organisation | Sectoral classification of organisation leading measures work. | get_measures
see_also | Additional details about the protected area specified. | get_pa
swmi | SWMI (Significant Water Management Issue) classification of the RNAG, for example "Diffuse source", "Invasive non-native species". | get_rnag
swmi_certainty | Assessment of the certainy with which the RNAG is linked to the SWMI classification specified. | get_rnag
title | Title of the measures project specified. | get_measures

